---
title: "Matrix Multiplication and Time Complexity"
author: "Daryl Adopo"
date: "9/5/2021"
output: pdf_document
---

```{r, message = FALSE, warning = FALSE}
library(microbenchmark)
library(tidyverse)
library(dasc2594)
library(patchwork)
```

<!-- Possibly show them a simple matrix multiplication algorithm  -->
<!-- and how the time complexity is calculated -->

# Understanding Matrix Multiplication Algorithm

The Matrix Multiplication of two matrices of size $n$ x $n$, $\mathbf{A}$ and $\mathbf{C}$.
Is an $n$ × $n$ matrix $\mathbf{C}$ where `C[i, j]` is the sum of the values of the ith row of $\mathbf{A}$ and the jth column of $\mathbf{B}$ multiplied together. 
<!-- They don't know dot products yet. -->
The elementary algorithm for matrix multiplication can be implemented as three nested loops.

```{r}
# use set.seed for reproduciblity
set.seed(2021)

# Iterative Approach
# A %*% B
matrix_multiply <- function(A, B){
  n = ncol(B)
  C <- matrix(0, n, n)
  for(i in 1:n){
    for(j in 1:n){
      for(k in 1:n){
        C[i, j] <- C[i, j] + A[i, k] * B[k, j]
      }
    }
  }
  return(C)
} 


n <- 2
A <- matrix(sample(1:10, n*2, replace = TRUE), n, n)
B <- matrix(sample(1:10, n*2, replace = TRUE), n, n)

# Checking our algorithm
all.equal(A %*% B, matrix_multiply(A, B))

bm <- microbenchmark(A %*% B, matrix_multiply(A, B))
bm

# Plotting the results using ggplot function autoplot
autoplot(bm)
```

Of course, performing the multiplication with the built-in operator `%*%` is way faster than our basic algorithm since it has been optimized through various techniques.

# Time Complexity

The time complexity of the matrix multiplication algorythm is calculated by summing the number 
of multiplications in the three nested loops.
\begin{align}
M(n,n,n) &= \sum_{i=1}^n\sum_{j=1}^n\sum_{k=1}^n 1 \\
 &= \sum_{i=1}^n\sum_{j=1}^n n \\
 &= \sum_{i=1}^n n^2 \\
 &= n^3
\end{align}

Therefore, will expect the mean processing time of our algorithm as $n$ increases to look like a cubic distribution.

Let's plot $n$ against the mean processing time for both the simple algorithm we created
and using `%*%`

```{r, warning = FALSE}
calc_time <- function(n, algo = TRUE) {
        ## create the matrices
        A <- matrix(rnorm(n^2), n, n)
        B <- matrix(rnorm(n^2), n, n)
        #calculate the time
        if(algo){
          bm <- microbenchmark(matrix_multiply(A, B))

        } else{
          bm <- microbenchmark(A %*% B)
          }
        # return the mean time
        return(mean(bm$time))
        
}
```

```{r, warning = FALSE}
n <- 1:10
# initialize the output vector
out <- length(n)

for (i in 1:length(n)) {
        out[i] <- calc_time(n[i])
}

dat <- data.frame(
        n = 1:10,
        mean_time = out)

plot1 <- dat %>% 
        ggplot(aes(x = n, y = mean_time)) +
        geom_point() +
        stat_smooth(method = "lm", color = "blue", se = FALSE) +
        stat_smooth(method = "lm", color = "red", formula = y ~ poly(x, 2), se = FALSE) +
        stat_smooth(method = "lm", color = "orange", formula = y ~ poly(x, 3), se = FALSE) +
        ggtitle("Matrix Multiply")

for (i in 1:length(n)) {
        out[i] <- calc_time(n[i], algo = FALSE)
}

dat <- data.frame(
        n = 1:10,
        mean_time = out)

plot2 <- dat %>% 
        ggplot(aes(x = n, y = mean_time)) +
        geom_point() +
        stat_smooth(method = "lm", color = "blue", se = FALSE) +
        stat_smooth(method = "lm", color = "red", formula = y ~ poly(x, 2), se = FALSE) +
        stat_smooth(method = "lm", color = "orange", formula = y ~ poly(x, 3), se = FALSE) +
        ggtitle("A %*% B")

plot1 + plot2
```

Although `A %*% B` is way faster than our `matrix_multiply` function, their processing time both follow a cubic function as $n$ increases. Thus, we say that matrix multiplication of $n \times n$ matrices is $O(n^3)$

